The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-08-24 conf_article/neon.tex
Article has been done

In this article, we explore the use of matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm that underlies the Muon update, we leverage the Ky Fan k-norms to introduce a new family of Muon-like algorithms we name {\it Fanions}. Then, we consider their convex combinations with the Frobenius norm and corresponding duals to develop a new family of algorithms we name {\it F-Fanions}, one of them being {\it F-Muon}. We complement our theoretical analysis with an extensive empirical study of the algorithms across a wide range of tasks and settings, from which it follows that properly-tuned F-Muon is on par with Muon, which raises the question about Muon's optimality.

@@

\section{Introduction}
 Minimizing loss functions in unprecedentedly high-dimensional spaces has recently become an integral and crucial part in training large language models. Hence, new scalable, time- and memory-efficient algorithms have been demanded. Besides well-known Adam \citep{kingma2014adam} and AdamW \citep{loshchilov2017decoupled}, recently proposed Muon has shown promising results on training very large models \citep{jordan2024muon,liu2025muon}. Its key difference from Adam and AdamW is that it has been constructed specifically for optimizing functions of weight matrices, which are common in deep learning.

@@

 In this article, we focus on the two most common matrix norms akin to the spectral, namely, the nuclear Norm and the Frobenius norm. Working in the linear minimization oracle (lmo) framework, which is equivalent to a factor to the trust region approach and the steepest descent under norm constraint, we derive Neon, our algorithm based on the nuclear norm. In the section {\it Matrix side of the updates}, we explain how Neon updates can be computed asymptotically faster than Muon updates by the Newton-Schulz iterations.

 Noticing that Neon and Muon are diametrical in terms of the rank of the update matrix, we bridge the space by "regularizing" them by NormalizedSGD, which is derived in lmo with the Frobenius norm. We do this in the same lmo approach by considering a norm that is dual to the convex combination of the Frobenius norm and the spectral or the nuclear norms respectively. So we derive the algorithms we name F-Neon and F-Muon respectively.

 Having faced the array of different Muon-like optimizers, according to the upper bounds from \citet{kovalev2025understanding,riabinin2025gluon}, with similar convergence behavior, we painstakingly compare them on a synthetic linear least squares problem with known Lipschitz constant. The efforts results in comparison of the algorithms by their convergence not in terms of the dual norms of their gradient, but in terms of the common spectral norm, which may strongly differ, especially in large matrices, from the initial dual norms. Thus, we compare the algorithms in a unified fashion.

 Finally, we test Muon, Neon, NSGD, F-Muon and F-Neon on deep-learning tasks: training convolutional network on CIFAR-10 and fine-tuning NanoGPT. The results support the supremacy of Muon, but the most striking result of the tests is that F-Muon, only half of which is Muon, surpasses Muon's accuracy on the CIFAR tasks by a margin. The case of F-Muon answers in the affirmative to the question of feasibility of constructing a mixture of optimization algorithms to increase robustness of the composite algorithm.

@@

Per-cycle complexity is $O(mn^2 + n^2k + nk^2)$, where m and n are dimensions of target matrix and n is the smallest one, k is retained subspace's size.

@@

During the research it was noted that RSVD can give good and fast approximation for singular values, but the matrix of approximation is far from the one given by truncated SVD, while TRLan gives good and fast approximation of a matrix, but not so good approximation for singular values. That means that TRLan may be not a perfect choice for algorithms like Dion, where $\sigma_i$ are required for error feedback.
\begin{table}[h!]
    \centering
    \begin{tabular}{c|c|c|c}
    \hline
    Method                              & rtol        & k & time,s \\
    \hline 
    Power Iterations                    & 0.01        & 1 & 7.7\\ 
    SVDS (thick-restart Lanczos method) & 0.01        & 1 & 0.18\\
    PCA Low Rank (RSVD)                 & 0.01        & 1 & 1.15\\
    SVDS (thick-restart Lanczos method) & 0.01        & 10 & 0.47\\
    PCA Low Rank (RSVD)                 & 0.01        & 10 & 19.4\\
    SVDS (thick-restart Lanczos method) & 0.01        & 100 & 1.96\\
    PCA Low Rank (RSVD)                 & 0.01        & 100 & 170\\
    \end{tabular}
    \caption{k-rank updated comparison}
    Comparison of different numerical methods to calculate k-rank update on $5000\times5000$ matrix of real numbers, rtol is an error in Frobenius norm relative to the k-rank approximation. 
    \label{tbl:matrix_methods}
\end{table}
\section{Trust Region Bounds for L-smooth functions}
    First, we analyze the problem in the unstochastic case. From Corollary 1 of \citet{kovalev2025understanding}, we directly get the following result that matches lower bounds, as was noted in that article.

@@

\section{Experiments}
    \iffalse
    \subsection{Randomized Linear Least Squares}
    Since the provided by other authors \citep{kovalev2025understanding,riabinin2025gluon} theoretical guarantees are almost norm-independent, we have to test them in practice.

    To test the bounds from \citet{kovalev2025understanding} in practice, we construct the following L-smooth problem:

@@

    We are interested to get $\epsilon_2=$ 1e-1 precision in $\norms{\cdot}$. Hence, we target $\epsilon = \epsilon_2$ for the nuclear norm (Neon), $\epsilon = \frac{n}{2}$ for the spectral norm (Muon), and $\epsilon = \frac{\sqrt{n}}{2}$ for the Frobenius norm (NSGD).

    {\bf An independent experiment}
    To make the problem more real-world, we do not use known information about smoothness. We run NormalizedSGD, Muon, F-Muon with $\alpha=1/2$, Neon, and F-Neon with $\alpha=1/2$ for 100 000 iterations with learning rate = 1e-3.

    \subsection{Logistic Regression}
    \fi
    \subsection{CIFAR-10 airbench}
    We adapt Keller Jordan's code to test F-Muon, Neon, and F-Neon on the CIFAR-10 airbench \citep{cifar2023airbench}.
    First, we run F-Muon for different $\alpha$ with the same {\tt lr=0.24(1 - step/total\_steps), momentum=0.6, nesterov=True}, as have been finetuned by Jordan. We record the accuracy after 8 epochs of training.

    % add figure

    Then, we tune F-Muon with $\alpha = 0.5$. Tuned parameters are {\tt lr=0.4(1 - step/total\_steps), momentum=0.65, nesterov = True}. While Muon reaches $94.00 \pm 0.13$\% accuracy after 8 epochs, tuned F-Muon reaches $94.01 \pm 0.14\%$ after 8 epochs.

    Finally, we take the set of tuned parameters and test on different $\alpha$. We notice that even when $\alpha=0.1$, the accuracy is much higher than in case of the pure NSGD.

@@

    \subsection{Modded NanoGPT speedrun}
    We test F-Muon on NanoGPT speedrun \citep{modded_nanogpt_2024}. For $\alpha = 0.5$, the optimal parameters are {\tt lr=0.07, momentum=0.95}, while for Muon they were  {\tt lr=0.05, momentum=0.95}. After testing for 1750 steps, as has been done on the speedruns, we get 3.281 cross-entropy loss, while on Muon, it falls below the target threshold 3.28 reaching 3.279.

    TO-DO: more discussions of experiments.

@@

\section{Formal assumptions}
Here we restate assumptions from \citet{kovalev2025understanding} that are used in the theorems from the main part of the article.
{\bf Lipschitz continuous gradient.}
    We assume that the gradient $\nabla f(\cdot)$ is Lipschitz continuous with respect to the norm $\norm{\cdot}$, that is, the following inequality holds:
    \begin{aequation}\label{eq:L}
    \norm{\nabla f(\mX) - \nabla f(\mX')}^\dagger \leq L \norm{\mX-\mX'}
    \quad\text{for all}\;
    \mX, \mX' \in \Rmn,
    \end{aequation}
    where $L > 0$ is the gradient Lipschitz constant.

>>>>

In this article, we explore the use of matrix norms for optimizing functions of weight matrices, a crucial problem in training large language models. Moving beyond the spectral norm that underlies the Muon update, we leverage the Ky Fan k-norms to introduce a new family of Muon-like algorithms we name {\it Fanions}. Then, we consider their convex combinations with the Frobenius norm and corresponding duals to develop a new family of algorithms we name {\it F-Fanions}, one of them being {\it F-Muon}. We complement our theoretical analysis with an extensive empirical study of the algorithms across a wide range of tasks and settings, from which it follows that properly-tuned F-Muon is almost on par with Muon, which questions the necessity of finding the exact solution to the problem in linear optimization oracle framework.

@@

\section{Introduction}
 Minimizing loss functions in unprecedentedly high-dimensional spaces has recently become an integral and crucial part in training large language models. Hence, new scalable, time- and memory-efficient algorithms have been demanded. Besides well-known Adam \citep{kingma2014adam} and AdamW \citep{loshchilov2017decoupled}, recently proposed Muon \citep{jordan2024muon} has shown promising results on training very large models \citep{liu2025muon}. Its key difference from Adam and AdamW is that it has been constructed specifically for optimizing functions of weight matrices, which are common in deep learning.

@@

 In this article, we focus on the two most common matrix norms akin to the spectral, namely, the nuclear norm and the Frobenius norm. Working in the linear minimization oracle (lmo) framework, which is equivalent to a factor to the trust region approach and the steepest descent under norm constraint, we derive Neon, our algorithm based on the nuclear norm. In the section {\it Matrix side of the updates}, we explain how Neon updates can be computed asymptotically with the same $\cO(m n \min\{m, n\})$ number of FLOPs as Muon updates are computed via Newton-Schulz iterations.

 Noticing that Neon and Muon are diametrical in terms of the rank of the update matrix, we bridge the space by "regularizing" them by Normalized SGD, which is derived in lmo with the Frobenius norm. We do this in the same lmo approach by considering a norm that is dual to the convex combination of the Frobenius norm and the spectral or the nuclear norm respectively. So we derive the algorithms we name F-Neon and F-Muon respectively.

 Having faced the array of different Muon-like optimizers, according to the upper bounds from \citet{kovalev2025understanding,riabinin2025gluon}, with similar convergence behavior, we painstakingly compare them on a synthetic linear least squares problem with known Lipschitz constant. The efforts results in comparison of the algorithms by their convergence not in terms of the dual norms of their gradient, but in terms of the common spectral norm, which may strongly differ, especially in large matrices, from the initial dual norms. Thus, we compare the algorithms in a unified fashion and find that if an algorithm converges fast by the loss, it does not mean it converges fast by the norm of the gradient, and vice versa.

 Finally, we test Muon and F-Muon on two deep-learning tasks, training convolutional network on CIFAR-10 \citep{cifar2023airbench} and pre-training NanoGPT on FineWeb \citep{modded_nanogpt_2024}. The results support the supremacy of Muon, but the most striking result of the tests is that F-Muon, only half of which is Muon, surpasses Muon's accuracy on the CIFAR tasks by a margin. The case of F-Muon answers in the affirmative to the question of feasibility of constructing a mixture of optimization algorithms to increase robustness of the composite algorithm.

@@

Per-cycle complexity is $O(mn^2 + n^2k + nk^2)$, where m and n are dimensions of the target matrix and n is the smaller one, k is retained subspace's size.

@@

During the research it was noted that RSVD can give good and fast approximation for singular values, but the matrix of approximation is far from the one given by truncated SVD, while TRLan gives good and fast approximation of a matrix, but not so good approximation for singular values. That means that TRLan may be not a perfect choice for algorithms like Dion, where $\sigma_i$ are required for error feedback.

The practical drawback of the TRLAn is the absence of its implementation in PyTorch.
\begin{table}[h!]
    \centering
    \begin{tabular}{c|c|c|c}
    \hline
    Method                              & rtol        & k & time,s \\
    \hline 
    Power Iterations                    & 0.01        & 1 & 7.7\\ 
    SVDS (thick-restart Lanczos method) & 0.01        & 1 & 0.18\\
    PCA Low Rank (RSVD)                 & 0.01        & 1 & 1.15\\
    SVDS (thick-restart Lanczos method) & 0.01        & 10 & 0.47\\
    PCA Low Rank (RSVD)                 & 0.01        & 10 & 19.4\\
    SVDS (thick-restart Lanczos method) & 0.01        & 100 & 1.96\\
    PCA Low Rank (RSVD)                 & 0.01        & 100 & 170\\
    \end{tabular}
    \caption{k-rank updated comparison}
    Comparison of different numerical methods to calculate k-rank update on $5000\times5000$ matrix of real numbers, rtol is an error in Frobenius norm relative to the k-rank approximation. 
    \label{tbl:matrix_methods}
\end{table}
\section{Trust Region Bounds for L-smooth functions}
    First, we analyze the problem in the unstochastic case. From Corollary 1 of \citet{kovalev2025understanding}, we directly get the following result that matches lower bounds, as was noted in that article.

@@

\section{Experiments}
    \subsection{Randomized Linear Least Squares}
    Since the provided by other authors \citep{kovalev2025understanding,riabinin2025gluon} theoretical guarantees are almost norm-independent, we have to test them in practice.

    To test the bounds from \citet{kovalev2025understanding}, we construct the following L-smooth problem:

@@

    We are interested to get $\epsilon_2=$ 1e0 precision in $\norms{\cdot}$. Hence, we target $\epsilon = \epsilon_2$ for the nuclear norm (Neon), $\epsilon = \frac{n}{2}$ for the spectral norm (Muon), and $\epsilon = \frac{\sqrt{n}}{2}$ for the Frobenius norm (NSGD).
    With that, we get the following lrs: Muon lr: 2.33E-04, Neon lr: 2.10e-02, Frobenius (NSGD) lr: 1.75e-02, and the following number of iterations: Muon iter: 68, Neon iter: 11919, Frobenius iter: 323. L-constants are 171.4, 47.7, and 16.2 respectively. As we will see, these results do not relate well to the practice.

    To make the problem more real-world, we do not use known information about smoothness. We run NormalizedSGD, Muon, F-Muon with $\alpha=1/2$, Neon, and F-Neon with $\alpha=1/2$ for 5 000 iterations with learning rate = 1e-2, which has been close to computed theoretically above. The results are in \cref{fig:lls}.

    Though Neon displays the worst convergence in terms of loss, it is a leader in terms of the norm of the gradient. This might signify that the known bounds are not sufficient even in the case of a convex optimization: lmo-based algorithms may converge well by these bounds, but be impractical like Neon.

    This experiment may also show that F-Muon may be more robust under general conditions than Muon.

    \begin{figure}[h!]
        \centering
        \begin{subfigure}[t]{0.48\linewidth}
            \centering
            \includegraphics[width=\linewidth]{figs/simple_lls/loss_vs_iteration_50x50.pdf}
            \caption{Loss}
            \label{fig:lls_loss}
        \end{subfigure}
        \hfill
        \begin{subfigure}[t]{0.48\linewidth}
            \centering
            \includegraphics[width=\linewidth]{figs/simple_lls/gradient_spectral_norm_vs_iteration_50x50.pdf}
            \caption{Spectral norm of the gradient}
            \label{fig:lls_grad_norm}
        \end{subfigure}
        \caption{Linear least squares problem for a 50 x 50 matrix}
        \label{fig:lls}

    \end{figure}

    \subsection{CIFAR-10 airbench}
    We adapt Keller Jordan's code to test F-Muon, Neon, and F-Neon on the CIFAR-10 airbench \citep{cifar2023airbench}.
    First, we run F-Muon for different $\alpha$ with the same {\tt lr=0.24(1 - step/total\_steps), momentum=0.6, nesterov=True}, as have been finetuned by Jordan, 10 repetitions for each $\alpha$. We record the accuracy after 8 epochs of training. The results are in \cref{fig:muon_alphas}.

    Then we tune F-Muon with $\alpha = 0.5$. Tuned parameters are {\tt lr=0.4(1 - step/total\_steps), momentum=0.65, nesterov = True}. While Muon reaches $94.00 \pm 0.13$\% accuracy after 8 epochs, tuned F-Muon reaches $94.01 \pm 0.14\%$ after 8 epochs (average was done by 200 iterations).

    Finally, we take this set of tuned parameters and test on different $\alpha$, 5 times for each $\alpha$. The results are in \cref{fig:fmuon_alphas}. We notice that even when $\alpha=0.1$, the accuracy is much higher than in case of the pure NSGD.

    \begin{figure}[h!]
        \centering
        \begin{subfigure}[b]{0.48\linewidth}
            \centering
            \includegraphics[width=\linewidth]{figs/muon_tuned_diff_alpha.pdf}
            \caption{With parameters tuned for Muon}
            \label{fig:muon_alphas}
        \end{subfigure}
        \hfill
        \begin{subfigure}[b]{0.48\linewidth}
            \centering
            \includegraphics[width=\linewidth]{figs/fmuon_tuned_diff_alpha.pdf}
            \caption{With parameters tuned for F-Muon}
            \label{fig:fmuon_alphas}
        \end{subfigure}
        \caption{Mean accuracies for different $\alpha$ of F-Muon.}
        \label{fig:diff_alphas}
    \end{figure}

@@

    The most pathological case is $\alpha < 0$, which corresponds to the lmo with an area that is not a norm ball! Despite this violation, the mixture of algorithms reaches almost the same accuracy as vanilla Muon.

    \subsection{Modded NanoGPT speedrun}
    We test F-Muon on NanoGPT speedrun \citep{modded_nanogpt_2024}. For $\alpha = 0.5$, the optimal parameters are {\tt lr=0.07, momentum=0.95}, while for Muon they were  {\tt lr=0.05, momentum=0.95}. After testing for 1750 steps, as it has been done on the speedruns, we get 3.281 cross-entropy loss, while on Muon, it falls below the target threshold 3.28 reaching 3.279.

    Again, as on the CIFAR airbench, if we set $\alpha=-0.1$, F* to which is not a norm, we get a 3.2818 loss, which is not higher than for $\alpha=0.3$, for example.

@@

\section{Formal assumptions}
Here we restate assumptions from \citet{kovalev2025understanding} that are used in the theorems from the main part of the article.

{\bf Lipschitz continuous gradient.}
    We assume that the gradient $\nabla f(\cdot)$ is Lipschitz continuous with respect to the norm $\norm{\cdot}$, that is, the following inequality holds:
    \begin{aequation}\label{eq:L}
    \norm{\nabla f(\mX) - \nabla f(\mX')}^\dagger \leq L \norm{\mX-\mX'}
    \quad\text{for all}\;
    \mX, \mX' \in \Rmn,
    \end{aequation}
    where $L > 0$ is the gradient Lipschitz constant.
